---
title: "hw3"
author: "Christopher Panlasigui"
date: "3/17/2021"
output:
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  comment = NA,
  results = "hide"
  )
library(tidyverse)
library(lubridate)
library(formatR)
library(nycflights13)
library(kableExtra)
```

## 5.5.2


1. Consider re-using your code from problem #4 of HW2.  Use `lubridate` and other tidyverse command to add a column to flights called `cal_dep` which is a datetime (dttm) variable representing the actual (_not_ scheduled) departure date and time. 


    ```{r, q_5_5_2_1}
flightsDated <- flights %>% 
  mutate(cal_dep = make_datetime(year, month, day, 
                                 dep_time %/% 100, ## hour
                                 dep_time %% 100)) %>% ## minute
  select(-year, -month, -day, -dep_time) %>% ## hide columns being replaced by cal_dep
  select(cal_dep, everything()) 
    ```


## --


2. Which plane `(tailnum)` has the worst on-time record? Define "on-time" as having `arr_delay \le 0`; the on-time record is then the proportion of flights which were on-time.  Also include in your summary the total number of lights in the tibble for each plane. 

  
    ```{r, on-time}
flightsDated %>%
  drop_na(arr_delay) %>%
  mutate(arr_delayed = if_else(arr_delay <= 0, FALSE, TRUE)) %>%
  group_by(tailnum) %>%
  summarize(prop_delayed = mean(arr_delayed),
            ave_delay = mean(arr_delay, na.rm = TRUE),
            number_flights = n()) %>%
  ungroup() %>%
  arrange(desc(prop_delayed))
    ```


    (a) What do you notice about the planes with a 0% on-time record?
    
    ```{r, zero-on-time}
    flightsDated %>%
      drop_na(arr_delay) %>%
      mutate(arr_delayed = if_else(arr_delay <= 0, FALSE, TRUE)) %>%
      group_by(tailnum) %>%
      summarize(prop_delayed = mean(arr_delayed),
                ave_delay = mean(arr_delay, na.rm = TRUE),
                number_flights = n()) %>%
      ungroup() %>% filter(prop_delayed == 1) %>% 
      arrange(desc(number_flights))
    ```
    
    
      The planes with 0% on-time record have very few flights.  The highest number of flights with 0% on-time record is 5.
     
    
    (b) Which plane, among those with at least twenty flights, has the worst on-time record?
    
    
    ```{r, worst-on-time20}
    worstOnTime <- flightsDated %>% 
      drop_na(arr_delay) %>%
      mutate(arr_delayed = if_else(arr_delay <= 0, FALSE, TRUE)) %>%
      group_by(tailnum) %>%
      summarize(prop_delayed = mean(arr_delayed),
                ave_delay = mean(arr_delay, na.rm = TRUE),
                number_flights = n()) %>% 
      ungroup() %>% 
      filter(number_flights >= 20) %>%  ## filter only with 20 or more flights 
      slice(which.max(prop_delayed))
    ```
    
    The plane with the worst on-time record that has at least 20 flights is `r worstOnTime$tailnum`.  It has a total of `r worstOnTime$number_flights` flights on record with `r worstOnTime$prop_delayed` proportion of its flights delayed averaging `r round(worstOnTime$ave_delay, 0)` minutes per flight. 
    
    
## 5.7.1


3. What time of day should you fly if you want to avoid delays as much as possible? 


```{r, avoid_delays}
flightsDelayByHour <- flightsDated %>% 
  drop_na(cal_dep) %>%
  group_by(hour) %>% 
  summarise(arr_delay = mean(arr_delay, na.rm = TRUE)) %>% 
  arrange(arr_delay) %>%
  ungroup() %>%
  filter(arr_delay <= 0)
  
```
      
It appears the best time/hour to avoid delay is between `r min(flightsDelayByHour$hour)` and `r max(flightsDelayByHour$hour)` in a 24-hour format time.

4. For each destination, compute the total minutes of delay.  For each flight, compute the proportion of the total delay for its destination  

```{r destination, message=FALSE}
# total minutes of delay for each destination
flightsDated %>%
  filter(arr_delay > 0) %>% # remove negative or zero delay
  group_by(dest) %>%
  summarise(total_arr_delay = sum(arr_delay))

# proportion of the total delay for each of flight's destination
flightsDated %>%
  filter(arr_delay > 0) %>% ## remove negative or zero delay
  group_by(dest, origin, carrier, flight) %>% 
  summarise(total_arr_delay = sum(arr_delay)) %>% 
  mutate(prop_delay = total_arr_delay/sum(total_arr_delay)) %>% 
  arrange(dest) %>% 
  select(-total_arr_delay)
```


## --

4. Find the five worst days of 2013 for delays.  Define the worst days as those with the highest proprotion of flights which are either cancelled or have departure delay over 15 minutes.  Then use the appropropriate join to add the `weather` data to `flights`.  Does extreme weather explain the delays on these days? 

```{r, worstdays, results='asis'}
weatherDated <- weather %>% 
  mutate(cal_dep = make_datetime(year, month, day)) %>% 
  select(-year, -month, -day, -hour) %>% 
  select(cal_dep, everything()) %>%
  group_by(day = floor_date(cal_dep, "day")) %>%
  summarise(ave_temp = mean(temp),
            ave_dewp = mean(dewp),
            ave_humid = mean(humid),
            ave_pressure = mean(pressure),
            ave_visib = mean(visib)) 


flightsCancelDelay <- flightsDated %>% 
  mutate(cancel_delay = if_else(dep_delay > 15 | is.na(cal_dep), TRUE, FALSE)) %>%
  select(cal_dep, origin, dep_delay, cancel_delay) %>% 
  #filter(cancel_delay == TRUE) %>%
  group_by(day = floor_date(cal_dep, "day")) %>%
  summarise(n_flights = n(),
            n_cancel_delay = sum(cancel_delay),
            prop_cancel_delay = mean(cancel_delay)) %>%
  drop_na(day) %>%
  arrange(desc(prop_cancel_delay)) %>% 
  inner_join(weatherDated, by = "day")

kbl(flightsCancelDelay %>% select(day, prop_cancel_delay, ave_humid) %>%
      head(n = 5))



```


The five worst days of 2013 delays might be due to high humidity.  


```{r, flightsModel_plot}
flightsModel <- lm(prop_cancel_delay ~ ave_humid, data = flightsCancelDelay)
summary(flightsModel)
plot(prop_cancel_delay ~ ave_humid, 
     data = flightsCancelDelay, 
     main = "Does Humidity Contribute To Delay or Cancellation of Flights?",
     xlab = "average humidity", 
     ylab = "proportion of cancelled or delayed flights")
abline(flightsModel, col = "red")
```

